Move Schedules: Fast persistence computations in sparse dynamic settings \thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-08-25 move_schedules.tex
Reorganized and editted section 3.3

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams given a time-varying filtration, and to detect homological critical events during the construction.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. The seminal work by Baik et al.~\cite{baik1999distribution} established that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c = -1.77108...$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. Since this lower bound grows similarly to $m!$ when $t \sim O(m)$ and $d << m$ fixed, we expect $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

\subsection{Minimizing schedule cost}\label{sec:schedule_cost}
\subsubsection{Motivation}
The results from the previous section yield a sufficient algorithm for generating move schedules of minimal cardinality: simply compute a $\mathrm{LCS}(K_0, K_1)$ for a pair $(K_0, K_1)$ and fix an order of simplices in the set $\mathrm{LCS}(K_0, K_1) \smallsetminus K_0$ to move. Any sequence of moves which increases the LCS on each move is guaranteed to have size $m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, and the reduction to the permutation edit distance problem ensures this size is optimal. 
However, like in the vineyards algorithm, certain pairs of simplices cost more to exchange depending on whether they are critical pairs in the sense described~\cite{cohen2006vines}, making specific move operations more expensive than others. 
%This variation in cost exaggerates the variability of the cost of scheduling move operations.

One advantage of Algorithm~\ref{alg:mr} is that the cost of applying a permutation to some given $R = D V$ decomposition can be determined efficiently prior to performing any column operations. 
Specifically, the cost of determining $\lvert \mathbb{I}\rvert + \lvert \mathbb{J}\rvert$ in lines (2) and (3) of Algorithm~\ref{alg:mr} cannot take more than $O(m)$ time. If we assume a similar complexity to determine the cost of applying a permutation with Algorithm~\ref{alg:ml} and we know to execute $d$ moves, an initial thought might be to use an optimization method like dynamic programming. 
That is, suppose $O(dm)$ move costs are computed ahead of time and these costs are stored in a table providing $O(1)$ element access and modification. Using a sparse row representation for $V$, this cannot take more than $O(dm)$ time, which is relatively efficient when $d << m$. With this structure, determining the cost of the cheapest move takes just $O(d)$ time. However, moving a simplex from position $i$ to position $j$ may require updating $O(\kappa^2)$ entries in the structure, where $\kappa = \lvert i - j \rvert$. Since $\kappa \sim O(m)$ in the worst case, direct optimization may require upwards of $\approx O(dm^2)$. Since this time complexity is potentially more expensive than executing a schedule to begin with, this direct strategy does not seem viable for practical settings.

@@

As before, given two permutations $p, q \in S_m$, let $S = ( s_1, s_2, \dots, s_d )$ denote a sorting such that $S \circ p = q$. 
In general, there are exponentially many ways to generate such a sorting $S$. 
The example given in section~\ref{sec:greedy} shows all (minimal size) $d!$ possible choices of $S$ for a fixed pair of permutations and choice of $\mathrm{LCS}(p,q)$. 
Let $\tau = \mathrm{LCS}(p,q)$ and let $\sigma \in S_d$ denote a choice of ordering   for the symbols in  $p \smallsetminus \tau$, 
such that $S(\sigma)$ denotes a schedule generated from this ordering. Ideally, we would like to achieve something similar to the following objective: 
\begin{equation}\label{eq:schedule_opt}
	S_\sigma^\ast = \argmin_{\sigma \in S_d} \mathrm{cost}(S(\sigma) \circ p)
where $\mathrm{cost}(S(\sigma) \circ p)$ measures the number of column operations require to execute $S(\sigma)$ on an existing decomposition. 

@@

When the set of permutations $S$ are adjacent transpositions, equality is achieved above since $K_\tau(\hat{s}_i, \hat{s}_{i+1}) = 1$ for every inversion between $p$ and $q$. However, this is not guaranteed if $S$ was derived using the strategy from section~\ref{sec:move_schedules}. Obtaining $S_\sigma^\ast$ via direct optimization over $S_d$ isn't practical, but experimental tests suggest that a greedy-type procedure performs well as a heuristic, which yields a $O(d^2 m \log m)$ minimization procedure. When $d <<  m$, this strategy of successively minimizing the Kendall-$\tau$ distance is efficient. 

@@

Although very similarity to the Kendall-$\tau$ distance, the Spearman distance offers some convenient computational advantages: it can be computed in $O(m)$ time, and the variance of its asymptotic distribution is larger. Consider the following minimization: 

@@

The minimizer of this optimization can be interpreted as a sorting which minimizes the net displacement amongst all possible sortings, again in the $\ell_1$ sense. 

@@

Given two permutations $p,q \in S_m$ and a LCS $\tau = \mathrm{LCS}(p,q)$ between them, the analysis in section~\ref{sec:move_schedules} shows we must move each symbol $x \in \mathcal{D}$ where $\mathcal{D} = p \smallsetminus \tau$ in a way that extends $\tau$ after each move. Thus we have $d = \lvert \mathcal{D} \rvert$ symbols to choose from initially. Each symbol $x \in \mathcal{D}$ yields a permutation $s$ to apply to $p$. Unlike the Kendall distance, the Spearman distance satisfies: 
For any sorting $S$ such that $s_{i}$ that increases the size of longest common subsequence between $s_{i-1} \circ \dots \circ s_1 \circ p$ and $q$. Since $\lvert \mathcal{D} \rvert$ is fixed, we would like to choose $x \in \mathcal{D}$ such that the corresponding permutation $s_x$ reduces the total displacement of every symbol $x \in \mathcal{D}$ relative to  $q$. A greedy strategy chooses $x$ such that $F(\hat{s}_i, \hat{s}_{i+1})$ is as small as possible, removes that symbol from $\mathcal{D}$, and recurses. Thus, a greedy strategy depends on our ability to quickly compute $F(\hat{s}_i, \hat{s}_{i+1})$. 

@@

All three schedules were generated from the same $\mathrm{LCS}(p,q) = (4\,7\,8)$ and all three schedules transform $p \mapsto q$ in $d = 6$ moves. The greedy heuristic proposed to minimize equation~\ref{eq:schedule_opt_succ} was used to produce $S_1$, which has the minimal number of crossings amongst all possible schedules in this example, since $K_\tau(p,q) = 21$. This minimization problem is related to a constrained version of the bipartite crossing minimization problem for $k$ sets of permutations, which is NP-hard for $k \geq 4$~\cite{biedl2009complexity}.

@@

The monotonic nature of successively applying the Spearman distance enables certain computational advantages. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized to $0$. Summing the entries in this array yields the Spearman distance in $O(m)$ time, which at initialization represents $F(p, p)$. Each symbol $x \in \mathcal{D}$ induces a  permutation $s_x$ such that $F(p, q) \geq F(s_x \circ p, q)$. Since there are $d = \lvert \mathcal{D} \rvert$ symbols to move and computing $F$ takes $O(m)$ time, a na\"ive greedy strategy requires $O((d+(d-1)+(d-2)+\dots+1)m) \approx O(d^2 m)$ time to complete. However, observe that each permutation $s_x$ changes the displacement of every symbol in a very predictable way: if $s_x$ represents a move permutation that moves $x$ from position $i$ to position $j$, then relative to the previous permutation $s_{i}$, the displacement of $x$ in $s_{i+1}$ changes by at most  $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. All other symbols are unaffected. Since the Spearman distance is simply the sum of these displacements, replacing $\mathcal{A}$ with a structure that supports both $O(\log m)$ access time to aggregate information (such as the sum) and $O(\log m)$ modification time ability to up to $\lvert i - j \vert$ entries. 
The former problem reduces to the problem of efficiently calculating and updating \emph{prefix sums}, however since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$, it's not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time. Fortunately, since every elements displacement other than $x$ changes by $\pm 1$ per update, both queries can be efficiently done in $O(\log m)$ time using a Segment Tree that supports \emph{lazy propagation}~\cite{de1997computational}. 

@@

%This cumulative Spearman distance enables certain computational advantages. Suppose we begin with an array $\mathcal{D}$ of size $m$ which provides $O(1)$ access and modification, initialized to $0$, whose entries represent the summands of equation~\ref{eq:schedule_opt_succ}. 
%observe that each entry 
%
%%\begin{align}
%%	
%%\end{align}
%%$$ \argmin_{\sigma \in S_d} \frac{1}{2} \sum\limits_{i = 1}^{d - 1} F(\hat{s}_i, \hat{s}_{i+1}) = \argmin_{\sigma \in S_d} \frac{1}{2} \sum\limits_{i = 1}^{d - 1}  \sum\limits_{i \in [m]} \lvert \hat{s}(i) - \hat{s}_{i+1}(i) \rvert$$ 
%%
%% = 
%
%such that summing the entries in this array yields the Spearman distance in $O(m)$ time, which at initialization represents $F(p, p)$. Each symbol $x \in \mathcal{D}$ induces a  permutation $s_x$ such that $F(p, q) \geq F(s_x \circ p, q)$. Since there are $d = \lvert \mathcal{D} \rvert$ symbols to move and computing $F$ takes $O(m)$ time, a na\"ive greedy strategy requires $O((d+(d-1)+(d-2)+\dots+1)m) \approx O(d^2 m)$ time to complete. However, observe that each permutation $s_x$ changes the displacement of every symbol in a very predictable way: if $s_x$ represents a move permutation that moves $x$ from position $i$ to position $j$, then relative to the previous permutation $s_{i}$, the displacement of $x$ in $s_{i+1}$ changes by at most  $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. All other symbols are unaffected. Since the Spearman distance is simply the sum of these displacements, replacing $\mathcal{A}$ with a structure that supports both $O(\log m)$ access time to aggregate information (such as the sum) and $O(\log m)$ modification time ability to up to $\lvert i - j \vert$ entries. 
%The former problem reduces to the problem of efficiently calculating and updating \emph{prefix sums}, however since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$, it's not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time. Fortunately, since every elements displacement other than $x$ changes by $\pm 1$ per update, both queries can be efficiently done in $O(\log m)$ time using a Segment Tree that supports \emph{lazy propagation}~\cite{de1997computational}.

>>>>

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, and to detect homological critical events during the construction.

@@

\begin{corollary}\label{cor:expectation}
	If $K_0, K_1$ are random distinct filtrations each of size $m$, then $d$ is upper bounded by $O(m - \sqrt{m})$ in expectation, with probability $1$ as $m \to \infty$.
\end{corollary}
\begin{proof}
	The proof of this result reduces to showing the average length of the LIS for random permutations. Let $L(p) \in [1,m]$ denote the maximal length of a increasing subsequence of $p \in S_m$. 
	The essential quantity to show the expected length of $L(p)$ over all permutations: 
	$$ \ell_m = \mathbb{E} \, L(p) = \frac{1}{m!} \sum\limits_{p \in S_m} L(p)$$
	The history of the mathematics of estimating this quantity dates back at least 50 years, and is sometimes called the \emph{Ulam-Hammersley} problem. The seminal work by Baik et al.~\cite{baik1999distribution} established that as $m \to \infty$:
	$$ \displaystyle \ell_m = 2 \sqrt{m} + c m^{1/6} + o(m^{1/6}) $$
where $c = -1.77108...$. As a result, we have: 
$$ \frac{\ell_m}{\sqrt{m}} \to 2\rlap{\quad \text{as } $m \to \infty$} $$	 
Thus, if $p \in S_m$ denotes a uniformly random permutation in $S_m$, then $L(p)/\sqrt{m} \to 2$ in probability as $m \to \infty$. 
Since the schedule size is defined as $d = m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, the claimed bound follows.
\end{proof}
\noindent The upper bound from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ under worst-case conditions. It is worth noting that this bound applies to pairs of uniformly sampled permutations, as opposed to uniformly sampled distinct filtrations---intuitively, simplicial filtrations have more structure due to the subcomplex relation that must be respected in the filtration. Thus, while $O(m - \sqrt{m})$ in expectation is an improvement over $O(m)$, it is still not tight. However, Boissonnat~\cite{boissonnat2018efficient} has shown that for any $d$-dimensional simplicial complex $K$ with $m$ simplices, the number of distinct filtrations built from $K$ is \emph{at least} $\floor{\frac{t+1}{d+1}}^m$, where $t$ is the number of distinct filtration values associated with the complex $K$. In the case where $t \sim O(m)$ and $d << m$ fixed, this lower bound grows similarly to $m!$---thus, $O(m - \sqrt{m})$ is not too pessimistic of an upper bound when the pair of filtrations are truly random. In practice, when one has a time-varying filtration and the sampling points are relatively close [in time], the LCS between adjacent filtrations is expected to be much larger, subsequently making $d$ much smaller. 
For example, the average size of the LCS across the 10 filtrations from Section~\ref{sec:motivation} each having $m=417$ simplices sampled at evenly spaced time points was $343$, implying $d \approx 70$ permutations needed on average to update the decomposition between adjacent time points.

\subsection{Minimizing schedule cost}\label{sec:schedule_cost}
\subsubsection{Motivation}
The results from the previous section yield a sufficient algorithm for generating move schedules of minimal cardinality: simply compute a $\mathrm{LCS}(K_0, K_1)$ for a pair $(K_0, K_1)$ and fix an order of simplices in the set $\mathrm{LCS}(K_0, K_1) \smallsetminus K_0$ to move. Any sequence of moves which increases the LCS on each move is guaranteed to have size $m - \lvert \mathrm{LCS}(K_0, K_1) \rvert$, and the reduction to the permutation edit distance problem ensures this size is optimal. 
However, like in the vineyards algorithm, certain pairs of simplices cost more to exchange depending on whether they are critical pairs in the sense described~\cite{cohen2006vines}, making specific move operations more expensive than others. 
This small variation in cost-per-transposition induces a large variability in the cost of executing any given schedule, which is undesirable computationally. Ideally, we would like to generate a schedule which not only is minimal in size, but also cost. Formally, given two permutations $p, q \in S_m$, let $S = ( s_1, s_2, \dots, s_d )$ denote a sorting such that $S \circ p = q$. 
%The example given in section~\ref{sec:greedy} shows all (minimal size) $d!$ possible choices of $S$ for a fixed pair of permutations and choice of $\mathrm{LCS}(p,q)$. 
Let $\nu = \mathrm{LCS}(p,q)$ and let $\sigma \in S_d$ denote a choice of ordering for the symbols in  $p \smallsetminus \nu$, such that $S(\sigma)$ denotes a schedule generated from this ordering. Rephrasing from the description above, we would like to achieve something similar to the following objective: 
\begin{equation}\label{eq:schedule_opt}
	S_\sigma^\ast = \argmin_{\sigma \in S_d} \mathrm{cost}(S(\sigma) \circ p)
\end{equation}
where $\mathrm{cost}(S(\sigma) \circ p)$ measures the number of column operations require to execute $S(\sigma)$ on an existing decomposition. There are exponentially many ways to generate such a sorting $S$. 
% Dynamic programming discussion 
One advantage of Algorithm~\ref{alg:mr} is that the cost of applying a permutation to some given $R = D V$ decomposition can be determined efficiently prior to performing any column operations. 
Specifically, the cost of determining $\lvert \mathbb{I}\rvert + \lvert \mathbb{J}\rvert$ in lines (2) and (3) of Algorithm~\ref{alg:mr} cannot take more than $O(m)$ time. If we assume a similar complexity to determine the cost of applying a permutation with Algorithm~\ref{alg:ml} and we know to execute $d$ moves, one might expect be to use an optimization method like dynamic programming. 
That is, suppose $O(dm)$ move costs are computed ahead of time and these costs are stored in a table providing $O(1)$ element access and modification. Using a sparse row representation for $V$, this cannot take more than $O(dm)$ time, which is relatively efficient when $d << m$. With this structure, determining the cost of the cheapest move takes just $O(d)$ time. However, moving a simplex from position $i$ to position $j$ may require updating $O(\kappa^2)$ entries in the structure, where $\kappa = \lvert i - j \rvert$. Since $\kappa \sim O(m)$ in the worst case, direct optimization may require upwards of $\approx O(dm^2)$. Since this time complexity is potentially more expensive than executing a schedule to begin with, this direct strategy does not seem viable for practical settings.

@@

@@

When the set of permutations $S$ are adjacent transpositions, equality is achieved above since $K_\tau(\hat{s}_i, \hat{s}_{i+1}) = 1$ for every inversion between $p$ and $q$, though this not guaranteed if $S$ was derived using the strategy from section~\ref{sec:move_schedules}. Obtaining $S_\sigma^\ast$ via direct optimization over $S_d$ isn't practical, but experimental tests suggest that a greedy-type procedure performs well as a heuristic, which yields a $O(d^2 m \log m)$ minimization procedure, which is efficient when $d <<  m$.

@@

Although very similar to the Kendall-$\tau$ distance, the Spearman distance offers some convenient computational advantages: it can be computed in $O(m)$ time, the variance of its asymptotic distribution is larger than Kendalls for uniformly random permutations, and it satisfies a certain monotonicity property under LCS edit operations. 
Specifically, given two permutations $p,q \in S_m$ with $\nu = \mathrm{LCS}(p,q)$ between them, the analysis in section~\ref{sec:move_schedules} shows we must move each symbol $x \in \mathcal{D}$ where $\mathcal{D} = p \smallsetminus \nu $ in a way that extends $\nu$ after each move. Each symbol $x \in \mathcal{D}$ induces a permutation $s_x$ such that $F(p, q) \geq F(s_x \circ p, q)$. That is, unlike the Kendall distance, the Spearman distance satisfies: 
For any sorting $S$ such that $s_{i}$ that increases the size of longest common subsequence between $s_{i-1} \circ \dots \circ s_1 \circ p$ and $q$. Since the complexity of the move operations discussed in section~\ref{sec:moves} depend on the size of the movement, it is advantageous to construct a schedule which minimizes this movement. 
%Since $\lvert \mathcal{D} \rvert$ is fixed, we would like to choose $x \in \mathcal{D}$ such that the corresponding permutation $s_x$ reduces the total displacement of every symbol $x \in \mathcal{D}$ relative to  $q$. 
We consider the following objective modified from equation~\ref{eq:schedule_opt}: 

@@

The minimizer of this optimization can be interpreted as a sorting which minimizes the net displacement amongst all possible sortings, in the $\ell_1$ sense. This minimization problem is related to a constrained version of the bipartite crossing minimization problem for $k$ sets of permutations, which is NP-hard for $k \geq 4$~\cite{biedl2009complexity}. As a result, we instead consider a greedy approach. A greedy strategy requires $O((d+(d-1)+(d-2)+\dots+1)m) \approx O(d^2 m)$ time to complete, since there are $O(d)$ symbols to choose from and minimizing $F(\hat{s}_i, \hat{s}_{i+1})$ at each step takes $O(md)$. Since $\mathcal{D}$ is fixed, a greedy strategy thus depends on our ability to quickly compute $F(\hat{s}_i, \hat{s}_{i+1})$.  

@@

@@

All three schedules were generated from the same $\mathrm{LCS}(p,q) = (4\,7\,8)$ and all three schedules transform $p \mapsto q$ in $d = 6$ moves. The greedy heuristic proposed to minimize equation~\ref{eq:schedule_opt_succ} was used to produce $S_1$, which has the minimal number of crossings amongst all possible schedules in this example, since $K_\tau(p,q) = 21$. 

@@

The monotonic nature of successively applying the Spearman distance enables certain computational advantages. Suppose we begin with an array $\mathcal{A}$ of size $m$ which provides $O(1)$ access and modification, initialized with the (signed) \emph{displacement} of every element in $p$ to its corresponding position in $q$. Since the Spearman distance is simply the sum of these displacements, if we could replace $\mathcal{A}$ with a data structure that supports $O(\log m)$ access time to aggregate information (i.e. the sum) and $O(\log m)$ modification time ability on $\lvert i - j \vert$ entries (simultaneously), we would have an efficient $O(d \log m)$ greedy solution. 
The former problem reduces to the problem of efficiently calculating \emph{prefix sums}, which is easily solved. It's not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time, since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$. However, observe that each permutation $s_x$ changes the displacement of every symbol in a very predictable way: if $s_x$ represents a move permutation that moves $x$ from position $i$ to position $j$, then relative to the previous permutation $s_{i}$, the displacement of $x$ in $s_{i+1}$ changes by $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. All other symbol displacements are unchanged.  
%, and 
It is known that one can apply constant-factor updates to multiple values in a prefix sum in $O(\log m)$ time by any prefix-sum data structure that also supports \emph{range updates}, e.g. a Fenwick tree or a Segment Tree that supports lazy propagation~\cite{de1997computational}. Single element prefix sum modifications can also be achieved $O(\log m)$ time with these data structures, the time to compute $F(\hat{s}_i, \hat{s}_{i+1})$ is bounded above by $O(\log m)$ and thus the greedy solution to equation~\ref{eq:schedule_opt_succ} can be computed in $O(d \log m)$ time. 

@@

%This cumulative Spearman distance enables certain computational advantages. Suppose we begin with an array $\mathcal{D}$ of size $m$ which provides $O(1)$ access and modification, initialized to $0$, whose entries represent the summands of equation~\ref{eq:schedule_opt_succ}. 
%observe that each entry 
%%\begin{align}
%%	
%%\end{align}
%%$$ \argmin_{\sigma \in S_d} \frac{1}{2} \sum\limits_{i = 1}^{d - 1} F(\hat{s}_i, \hat{s}_{i+1}) = \argmin_{\sigma \in S_d} \frac{1}{2} \sum\limits_{i = 1}^{d - 1}  \sum\limits_{i \in [m]} \lvert \hat{s}(i) - \hat{s}_{i+1}(i) \rvert$$ 
%%
%% = 
%such that summing the entries in this array yields the Spearman distance in $O(m)$ time, which at initialization represents $F(p, p)$. Each symbol $x \in \mathcal{D}$ induces a  permutation $s_x$ such that $F(p, q) \geq F(s_x \circ p, q)$. Since there are $d = \lvert \mathcal{D} \rvert$ symbols to move and computing $F$ takes $O(m)$ time, a na\"ive greedy strategy requires $O((d+(d-1)+(d-2)+\dots+1)m) \approx O(d^2 m)$ time to complete. However, observe that each permutation $s_x$ changes the displacement of every symbol in a very predictable way: if $s_x$ represents a move permutation that moves $x$ from position $i$ to position $j$, then relative to the previous permutation $s_{i}$, the displacement of $x$ in $s_{i+1}$ changes by at most  $\lvert i - j \rvert$ and the displacement of any symbol between $i$ and $j$ changes by $\pm 1$. All other symbols are unaffected. Since the Spearman distance is simply the sum of these displacements, replacing $\mathcal{A}$ with a structure that supports both $O(\log m)$ access time to aggregate information (such as the sum) and $O(\log m)$ modification time ability to up to $\lvert i - j \vert$ entries. 
%The former problem reduces to the problem of efficiently calculating and updating \emph{prefix sums}, however since $\lvert i - j \rvert \leq m$ is potentially larger than $O(\log m)$, it's not immediately clear how to achieve the latter modification complexity in $O(\log m)$ time. Fortunately, since every elements displacement other than $x$ changes by $\pm 1$ per update, both queries can be efficiently done in $O(\log m)$ time using a Segment Tree that supports \emph{lazy propagation}~\cite{de1997computational}.
